---
title: "Antibiotics"
author: "Zach"
date: 2020-10-07
output:
  github_document
---

(Solution document)

*Purpose*: To create an effective visualization, we need to keep our *purpose* firmly in mind. There are many different ways to visualize data, and the only way we can judge efficacy is with respect to our purpose.

In this challenge you'll visualize the same data in two different ways, aimed at two different purposes.

*Note*: Please complete your initial visual design **alone**. Work on both of your graphs alone, and save a version to your repo *before* coming together with your team. This way you can all bring a diversity of ideas to the table!

<!-- include-rubric -->

```{r setup}
library(tidyverse)
library(ggrepel)
library(rzdr)
```

*Background*: The data[1] we study in this challenge report the [*minimum inhibitory concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration) (MIC) of three drugs for different bacteria. The smaller the MIC for a given drug and bacteria pair, the more practical the drug is for treating that particular bacteria. An MIC value of *at most* 0.1 is considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin, streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized into a genus based on a number of features, including their resistance to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()
```

# Visualization
<!-- -------------------------------------------------- -->

## Purpose: Compare Effectiveness
<!-- ------------------------- -->

__q1__ Create a visualization of `df_antibiotics` that helps you to compare the effectiveness of the three antibiotics across all the bacteria reported. Make sure to show *all* of the bacteria, all three antibiotic MIC values, and whether or not each bacterium is Gram positive/negative. Can you make any broad statements about antibiotic effectiveness?

```{r q1-task}
## TASK: Create your visualization
df_antibiotics %>%
  pivot_longer(
    names_to = "antibiotic",
    values_to = "mic",
    cols = c(penicillin, streptomycin, neomycin)
  ) %>%
  mutate(
    inverse_mic = 0.11 / mic,
    fct_reorder(antibiotic, inverse_mic)
  ) %>%

  ggplot(
    aes(fct_reorder(bacteria, inverse_mic), inverse_mic, fill = antibiotic)
  ) +
  geom_col(position = "dodge") +
  geom_vline(xintercept = seq(1.5, 10, 1), size = 0.2) +

  scale_y_log10(
    breaks = c(1e-4, 1e-3, 1e-2, 1e-1, 1e+0, 1e+1, 1e+2),
    labels = c("(Less Effective)", 0.001, 0.01, 0.1, "(Effective)", 10, "(More effective)")
  ) +
  coord_flip() +
  scale_fill_discrete(name = "Antibiotic") +
  facet_grid(gram~., scales = "free", space = "free", labeller = label_both) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title.position = "plot",
    panel.border = element_rect(fill = NA, size = 1),
    panel.background = element_rect(fill = "grey90"),
    panel.grid.major = element_line(color = "grey50"),
    axis.text.x = element_text(size = 14),
    axis.title.x = element_text(size = 16),
    axis.text.y = element_text(size = 14),
    axis.title.y = element_text(size = 16),
    strip.text.y = element_text(size = 14)
  ) +
  labs(
    x = "Bacteria",
    y = "0.1 / MIC",
    title = "Antibiotic Effectiveness Varies with Gram Stain",
    subtitle = "Penicillin and Streptomycin for Gram positive, Neomycin for (some) Gram negative"
  )
ggsave("./images/c05-antibiotics-all.png", width = 10, height = 6)
```

**Observations**:

- Among the antibiotics considered here, only Neomycin is effective for Gram negative bacteria
- Penicilin and Streptomycin tend to be more effective for Gram positive bacteria
- None of the bacteria here are effective for a large collection of Gram negative bacteria

**Visual design notes**

- The **purpose** of this graph is to compare all three antibiotic MIC values against a threshold. Therefore I pivoted the data in order to show all three MIC values on a single continuous axis.

- I used `geom_col()` because bars help emphasize a "base" number. I plot the mutated value `0.1 / MIC` to put that "base" at MIC = 0.1.
- I decided to use an axis to show the bacteria names, so I can show the Gram stain "for free" by using facets.
- Since there are a lot of bacteria, I found it effective to map the MIC values to the horizontal axis, so that the bacteria names can remain horizontal without overlap.
- This is a *highly subjective* point, but I find that "up" and "right" tend to correspond with "better" in visual language. Therefore my horizontal axis is actually *reversed*; I'm actually plotting `0.1 / MIC`. This effectively flips the axis to put more effective antibiotics towards the right of the plot.
- I order the factors by median MIC value to add some structure.
- I use manual `breaks` and `labels` in my `scale_` call to add some instructions on how to read this plot: note that the axis labels themselves tell the reader which direction is more or less effective.
- I draw attention to primary takeaways in the title and subtitle of the graph.

## Purpose: Categorize Bacteria
<!-- ------------------------- -->

The *genus* of a living organism is a human categorization, based on various characteristics of the organism. Since these categories are based on numerous factors, we will tend to see clusters if we visualize data according to relevant variables. We can use these visuals to categorize observations, and to question whether given categories are reasonable!

__q2__ Create a visualization of `df_antibiotics` that helps you to categorize bacteria according to the variables in the data. Here your emphasis **need not** be on whether MIC < 0.1; instead focus on detecting patterns between the individual bacteria.

Make sure to show *all* of the bacteria and whether or not each bacterium is Gram positive/negative; you do *not* need to show all three MIC values. Document your observations on how how clusters of bacteria in the variables do---or don't---align with their *genus* classification.

```{r q2-task}
## TASK: Create your visualization
df_antibiotics %>%
  mutate(genus = str_extract(bacteria, "^\\w+")) %>%

  ggplot(aes(neomycin, penicillin)) +
  geom_point(
    aes(shape = gram),
    size = 2.5
  ) +
  geom_point(
    data = . %>%
      filter(genus %in% c("Streptococcus", "Staphylococcus", "Salmonella")),
    mapping = aes(color = genus, shape = gram),
    size = 2.5
  ) +
  geom_text_repel(
    aes(label = bacteria),
    size = 3
  ) +
  scale_x_log10() +
  scale_y_log10() +
  theme_minimal()
ggsave("./images/c05-antibiotics-map.png")
```

**Observations**:

- *Diplococcus pneumonia* is near two species of *Streptococcus*, which is odd.
- *Streptococcus fecalis* is far from the other *Streptococcus* species, which is also odd.
- In 1974 *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*, and in 1984 *Streptococcus fecalis* was renamed *Enterococcus fecalis*.[2]

**Visual design notes**:

- The **purpose** of this graph is to look for relationships between bacteria. This is easiest to do if the plot is constructed such that the observations lie close to each other if they are related. Therefore, I choose to leave the data unpivoted, and plot in a two-dimensional space of `penicillin` vs `neomycin`.
  - In contrast with the graph above (where the observations are arbitrarily distant from each other, based on ordering factors), in this q2 graph observations naturally lie close or far from each other, since they are represented with points.
  
- I sub-select the genera for plotting, as it is far easier to distinguish between a smaller number of colors. Additionally there are many genera with only a single bacterium represented; those colors will essentially be wasted.
- Shapes are even more difficult to distinguish than colors; I map `gram` to `shape` because there are only two levels, which makes this easy to distinguish.
- I use a log-scale for both MIC values, as the values are very widely spread.

# References
<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." *American Scientist* (2009) [link](https://www.americanscientist.org/article/thats-funny)
