---
title: "Diamonds Report"
author: "Zachary del Rosario"
date: 2020-09-16
output:
  github_document
---

An example analysis of the `diamonds` dataset.

```{r setup}
library(tidyverse)
library(rzdr) # My custom library; https://github.com/zdelrosario/rzdr
library(cowplot)
```

# First Look
<!-- -------------------------------------------------- -->

You should **always** do your first checks when studying a dataset!

```{r glimpse}
diamonds %>% glimpse()
```

We have observations on `carat, cut, color, clarity, price`; these are the variables I focus on.

```{r summary}
diamonds %>% summary()
```

`summary` gives us a sense of variation across the entire dataset, but only in a univariate way.

# Carat seems most important
<!-- -------------------------------------------------- -->

Here's my figure that intimates that carat is the most important variable for
determining price.

```{r vis1}
diamonds %>%

  ggplot(aes(carat, price, color = cut)) +
  geom_smooth(se = FALSE) +

  scale_x_log10() +
  scale_y_log10() +
  viridis::scale_color_viridis(name = "Cut", discrete = TRUE) +
  theme_common() +
  labs(
    x = "Carat",
    y = "Diamond Price"
  )
```

Note that this is *not* conclusive by itself; other variables could
(conceivably) co-vary with carat and equally-explain the price. The regression
analysis below controls for other variables.

# Focus on pricing discontinuities
<!-- ------------------------- -->

The following are closer looks at the price-carat relationship, using boxplots.

```{r vis4}
p1.disco <-
  diamonds %>%
  filter(
    0.47 <= carat,
    carat <= 0.53
  ) %>%

  ggplot(aes(carat, price)) +
  geom_boxplot(aes(group = carat), notch = TRUE) +

  theme_common() +
  labs(
    x = "Carat",
    y = "Diamond Price"
  )

p2.disco <-
  diamonds %>%
  filter(
    0.67 <= carat,
    carat <= 0.73
  ) %>%

  ggplot(aes(carat, price)) +
  geom_boxplot(aes(group = carat), notch = TRUE) +

  theme_common() +
  labs(
    x = "Carat",
    y = "Diamond Price"
  )

plot_grid(
  p1.disco,
  p2.disco,
  nrow = 1,
  ncol = 2
)
```

Looking at these, I see the price jump around `carat == 0.5` and `carat == 0.7` as indicative of pricing anomalies. Essentially: People are willing to pay a lot more for diamonds that are at or above these "special" values.

# Model
<!-- -------------------------------------------------- -->

Fit model near "focal point" of `carat == 0.5` based on Scott and Yelowitz (2012).

```{r fit-focal-0.5}
fit_c0.5 <-
  diamonds %>%
  filter(
    0.4 <= carat,
    carat <= 0.6
  ) %>%
  mutate(bump = carat >= 0.5) %>%

  lm(
    data = .,
    formula = log(price) ~ carat + cut + color + clarity + bump
  )

summary(fit_c0.5)
```

These model results tell me a number of things:

- The regression coefficient on `carat` is *large*, indicating that this is a very important variable even when accounting for `cut, color, clarity` in the regression
  - This is called *controlling* for other factors (here `cut, color, clarity`)
- The `bump` variable is also *large*; this model term represents the "price bump" that occurs simply by having a carat at or above the target focal point of `carat == 0.5`. This is the "discontinuous" part of the regression.
  - This `bump` idea is related to the notion of a [regression discontinuity](https://en.wikipedia.org/wiki/Regression_discontinuity_design), which we *will not* cover in this course
