---
title: "Visual Tricks with Gapminder"
author: "Zach del Rosario"
date: 2020-07-25
output:
  github_document:
    toc: true
---

*Purpose*: This is an illustration of visual tricks related to `c04-gapminder`.

```{r setup}
library(tidyverse)
library(gapminder)
library(ggrepel)

## Setup related to the challenge
year_max <-
  gapminder %>%
  pull(year) %>%
  max()
year_min <-
  gapminder %>%
  pull(year) %>%
  min()
```

# Role of Scale Transforms
<!-- -------------------------------------------------- -->

Compare the following two graphs

```{r scales-natural, fig.height=4}
set.seed(101)

gapminder %>%
  filter(year == year_min) %>%

  ggplot(aes(fct_reorder(continent, gdpPercap), gdpPercap)) +
  geom_point(position = position_jitter(height = 0, width = 0.1)) +
  labs(
    x = "Continent",
    y = "GDP per Capita (USD)",
    caption = str_c("in ", year_min),
    title = "Scale: Natural"
  )
```

On a natural scale (non-transformed), the outlier in Asia drastically stretches
our scale, making it difficult to see variation among the other countries.

```{r scales-log10, fig.height=4}
set.seed(101)

gapminder %>%
  filter(year == year_min) %>%

  ggplot(aes(fct_reorder(continent, gdpPercap), gdpPercap)) +
  geom_point(position = position_jitter(height = 0, width = 0.1)) +
  scale_y_log10() +
  labs(
    x = "Continent",
    y = "GDP per Capita (USD)",
    caption = str_c("in ", year_min),
    title = "Scale: Log10"
  )
```

On a log-scale our visual range is compressed, making it easier to distinguish
the individual countries.

```{r}
gapminder %>%
  ggplot(aes(log10(gdpPercap))) +
  geom_histogram()
```

```{r}
gapminder %>%
  ggplot(aes(lifeExp)) +
  geom_histogram()
```

```{r}
gapminder %>%
  ggplot(aes(lifeExp, gdpPercap)) +
  geom_point()  +
  scale_y_log10()
```


**Remember that scales are a "knob" you should turn**

# Ordering of factors
<!-- -------------------------------------------------- -->

## MPG dataset
<!-- ------------------------- -->

*Ordering* our factor levels is very important when there are many levels:

Imagine trying to (quickly) tell which manufacturer has the generally
most-efficient vehicles in the dataset:

```{r mpg-unordered, fig.height=4}
mpg %>%

  ggplot(aes(manufacturer, cty)) +
  geom_boxplot() +
  coord_flip()
```

Try answering the same question with the following figure.

```{r mpg-ordered, fig.height=4}
mpg %>%
  mutate(manufacturer = fct_reorder(manufacturer, cty)) %>%

  ggplot(aes(manufacturer, cty)) +
  geom_boxplot() +
  coord_flip()
```

All that's changed between the two is the ordering of the *factor levels*.
Generally alphabetical order is meaningless.

## Gapminder Dataset
<!-- ------------------------- -->

Ordering factors still useful with the gapminder set.

```{r gapminder-unordered, fig.height=4}
gapminder %>%
  filter(year %in% c(year_max, year_min)) %>%

  ggplot(aes(continent, gdpPercap)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("Kuwait", "United States", "Canada")),
    mapping = aes(color = country)
  ) +
  scale_y_log10() +
  scale_color_discrete(name = "Outliers") +
  facet_grid(~year) +
  labs(
    x = "Continent",
    y = "GDP per Capita (USD)"
  )
```

Ordering the factors makes the graph easier to parse.

```{r gapminder-ordered, fig.height=4}
gapminder %>%
  filter(year %in% c(year_max, year_min)) %>%

  ggplot(aes(fct_reorder(continent, gdpPercap), gdpPercap)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("Kuwait", "United States", "Canada")),
    mapping = aes(color = country)
  ) +
  scale_y_log10() +
  scale_color_discrete(name = "Outliers") +
  facet_grid(~year) +
  labs(
    x = "Continent",
    y = "GDP per Capita (USD)"
  )
```

**If your factors do not have a meaningful order, use another variable to give them a meaningful order.**

# Use of Layers
<!-- -------------------------------------------------- -->

Trying to color-label all countries is a mess:

```{r gapminder-terrible, fig.height=4}
gapminder %>%
  filter(year %in% c(year_max, year_min)) %>%

  ggplot(aes(continent, gdpPercap)) +
  geom_point(
    mapping = aes(color = country)
  ) +
  scale_y_log10() +
  scale_color_discrete(name = "Outliers") +
  facet_grid(~year) +
  labs(
    x = "Continent",
    y = "GDP per Capita (USD)"
  )
```

We can do a *little* better by using text labels:

```{r gapminder-still-terrible, fig.height=4}
gapminder %>%
  filter(year %in% c(year_max, year_min)) %>%

  ggplot(aes(continent, gdpPercap)) +
  geom_point() +
  geom_text_repel(
    aes(label = country)
  ) +
  scale_y_log10() +
  scale_color_discrete(name = "Outliers") +
  facet_grid(~year) +
  labs(
    x = "Continent",
    y = "GDP per Capita (USD)"
  )
```

At least we can see something! Filtering at the *top* level removes a lot of the
context we're seeking:

```{r gapminder-filtered, fig.height=4}
gapminder %>%
  filter(year %in% c(year_max, year_min)) %>%
  group_by(year) %>%
  filter(dense_rank(desc(gdpPercap)) <= 5) %>%

  ggplot(aes(continent, gdpPercap)) +
  geom_point() +
  geom_text_repel(
    aes(label = country)
  ) +
  scale_y_log10() +
  scale_color_discrete(name = "Outliers") +
  facet_grid(~year) +
  labs(
    x = "Continent",
    y = "GDP per Capita (USD)"
  )
```

We would do better to use the *layers* functionality of ggplot; filter out the
data *within a specific layer* to emphasize points:

```{r gapminder-layer-filter, fig.height=4}
gapminder %>%
  filter(year %in% c(year_max, year_min)) %>%

  ggplot(aes(fct_reorder(continent, gdpPercap), gdpPercap)) +
  ## Boxplot gets all of the data
  geom_boxplot() +
  geom_point(
    ## NOTE: The filter *inside* one of the layers
    data = . %>% filter(country %in% c("Kuwait", "United States", "Canada")),
    mapping = aes(color = country)
  ) +
  scale_y_log10() +
  scale_color_discrete(name = "Outliers") +
  facet_grid(~year) +
  labs(
    x = "Continent",
    y = "GDP per Capita (USD)"
  )
```

**Use ggplot's layers to highlight important observations.**

# Permuting variables
<!-- -------------------------------------------------- -->

*Small changes* to our aesthetic assignments can have a large impact on
readability. For instance: The following two visuals are virtually identical
*except* that I've swapped two of the aesthetics: `continent` and `year`.

Compare version 1:

```{r gapminder-permutation1, fig.height=4}
gapminder %>%
  filter(year %in% c(year_max, year_min)) %>%

  ggplot(aes(fct_reorder(continent, gdpPercap), gdpPercap)) +
  geom_boxplot() +
  geom_point(
    ## NOTE: The filter *inside* one of the layers
    data = . %>% filter(country %in% c("Kuwait", "United States", "Canada")),
    mapping = aes(color = country)
  ) +
  scale_y_log10() +
  scale_color_discrete(name = "Outliers") +
  facet_grid(~year) +
  labs(
    x = "Continent",
    y = "GDP per Capita (USD)"
  )
```

Note that version 1 makes comparisons *within* a year *between* continents easy.

Compare with version 2:

```{r gapminder-permutation2, fig.height=4}
gapminder %>%
  filter(year %in% c(year_max, year_min)) %>%

  ggplot(aes(fct_reorder(as.factor(year), gdpPercap), gdpPercap)) +
  geom_boxplot() +
  geom_point(
    ## NOTE: The filter *inside* one of the layers
    data = . %>% filter(country %in% c("Kuwait", "United States", "Canada")),
    mapping = aes(color = country)
  ) +
  scale_y_log10() +
  scale_color_discrete(name = "Outliers") +
  facet_grid(~continent) +
  labs(
    x = "Year",
    y = "GDP per Capita (USD)"
  )
```

Note that the second one makes comparisons *within* a continent *between* years
easy.

Which version is "correct" *depends on the question you are trying to answer*.

**Design your visuals based on the comparison you are trying to make.**

"... the heart of quantitative thinking: “Compared to what?”" - Edward Tufte
