---
title: "Judging Trends Under Noise"
author: "Zach"
date: 2020-09-30
output:
  github_document
---

This short notebook is about judging trends when those trends are corrupted by noise.

```{r setup}
library(tidyverse)

filename <- "./data/stang.csv"
df_stang <- read_csv(filename)

df_stang_long <-
  df_stang %>%
  pivot_longer(
    names_to = c(".value", "angle"),
    names_sep = "_",
    cols = c(-thick, -alloy)
  ) %>%
  mutate(angle = as.integer(angle)) %>%
  filter(E >= 0, nu >= 0)
```

## Little data, a dubious trend
-------------------------

First, let's look at the elasticity against sample thickness:

```{r trend1}
df_stang_long %>%

  ggplot(aes(thick, E)) +
  geom_point(size = 3) +
  geom_smooth(se = FALSE) +
  rzdr::theme_common()
```

**Observations**:

- There is a general negative trend between `E` and `thick`
- However, note that this trend isn't *monotone*; it doesn't go *only down*, it goes down-up-down
  - For variables like elasticity and thickness, it's *dubious* that we would see a non-monotone trend like this. Therefore, I think this pattern is probably spurious.

## Lots of data, lots of noise
-------------------------

Next, let's look at a case where we have *tons* of data:

```{r trend2}
n <- 1e3
tibble(x = seq(0, 1, length.out = n)) %>%
  mutate(y = x + rnorm(sd = 3, n = n)) %>%

  ggplot(aes(x, y)) +
  geom_point()
```

**Observations**:

- Since I generated the data, *I know* there's a trend underlying the data.
- However, since there's *a ton* of noise, even having `1000` data points doesn't make things clearer.

## Little data, little noise
-------------------------

One last example:

```{r trend3}
n <- 15
tibble(x = seq(0, 1, length.out = n)) %>%
  mutate(y = sin(2 * pi * x) + rnorm(sd = 0.1, n = n)) %>%

  ggplot(aes(x, y)) +
  geom_point()
```

**Observations**:

- Since I generated the data, *I know* there's a trend underlying the data.
- Here we have just 15 observations (with Stang we had 26), but the trend is fairly clear.

# Punchline
--------------------------------------------------

"More data" is **not** always the right answer. It depends on:

- The strength of the underlying **signal** (or **trend**)
- The degree of **noise** lying on top of that signal
