---
title: "Challenges of visual comparison"
author: "Zach"
date: 2020-09-22
output:
  github_document
---

This notebook illustrates how difficult it is to differentiate between a power law

$$y = x^a$$

and an exponential

$$y = \exp(x)$$

through visual comparison alone.

```{r setup}
library(tidyverse)
```

Generate data from an exponential and a power law.

```{r gen-data}
df_data <-
tibble(x = seq(0, 10, length.out = 100)) %>%
  mutate(
    y_line = 1000 * x,
    y_exp = exp(x),
    y_pow = x ^ 4.5
  ) %>%
  pivot_longer(
    names_to = c(".value", "fun"),
    names_sep = "_",
    cols = c(y_line, y_exp, y_pow)
  ) %>%
  mutate(fun = fct_relevel(fun, "line", "exp", "pow"))
```

## Natural scale
<!-- ------------------------- -->

Can you visually distinguish between the two nonlinear functions?

```{r scale-natural}
df_data %>%
  ggplot(aes(x, y)) +
  geom_point() +

  facet_wrap(~fun) +
  theme_minimal()
```

## Semilog scale
<!-- ------------------------- -->

A semilog scale can help us see an exponential trend.

```{r scale-semilog}
df_data %>%
  ggplot(aes(x, y)) +
  geom_point() +

  scale_y_log10() +
  facet_wrap(~fun) +
  theme_minimal()
```

## Loglog scale
<!-- ------------------------- -->

A loglog scale can help us see a power law trend.

```{r scale-loglog}
df_data %>%
  ggplot(aes(x, y)) +
  geom_point() +

  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~fun) +
  theme_minimal()
```
