---
title: "RMS Titanic: Fishy Plots"
author: "Zach"
date: 2020-09-24
output:
  github_document:
    toc: true
---

```{r setup}
library(tidyverse)
df_titanic <- as_tibble(Titanic)
```

# Fishy Bar Charts
<!-- -------------------------------------------------- -->

Let's compute the proportions, as in the challenge

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()

df_prop
```

If we try to plot this with only aesthetics for `Sex, Class`, we get the following:

```{r q4-mistake}
df_prop %>%
  filter(Survived == "Yes") %>%

  ggplot(aes(Class, Prop, fill = Sex)) +
  geom_col(position = "dodge") +

  scale_y_continuous(labels = scales::label_percent())
```

Note that this *falsely* implies that everyone in the 1st and 2nd classes survived! Adding an aesthetic for `Age` helps reveal what went wrong:

```{r q4-fixed}
df_prop %>%
  filter(Survived == "Yes") %>%

  ggplot(aes(Class, Prop, fill = Sex, color = Age)) +
  geom_col(position = "dodge") +

  scale_y_continuous(labels = scales::label_percent())
```

Ggplot essentially hid some of the bars, which contributed to a false impression of the data. It turns out to not be a huge deal for the Female passengers (Adults and Children had not-too-dissimilar survival rates), but for Males this is *extremely misleading*.

**Punchline**: Know thine data! If there are variables you're not plotting, make sure you're at least **aware** of the fact that you're not plotting them!
